---
title: "The pcaMethods Package"
author:
  - name: Wolfram Stacklies
    affiliation: CAS-MPG Partner Institute for Computational Biology (PICB), Shanghai, P.R. China
  - name: Henning Redestig
    affiliation: Max Planck Institute for Molecular Plant Physiology, Potsdam, Germany
    url: http://bioinformatics.mpimp-golm.mpg.de/
date: "17 May 2026"
package: pcaMethods
output: BiocStyle::html_document
bibliography: pcaMethods.bib
reference-section-title: References
vignette: |
    %\VignetteIndexEntry{pcaMethods}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
---

# Overview

The `pcaMethods` package [@stacklies07] provides a set of
different PCA implementations, together with tools for cross
validation and visualisation of the results.  The methods basically
allow to perform PCA on incomplete data and thus may also be used for
missing value estimation.

When doing PCA one assumes that the data is restricted to a subspace
of lower dimensionality, e.g. correlation patterns between jointly regulated
genes. PCA aims to extract these structures thereby filtering noise out.
If only the most significant loadings (eigenvectors, also referred to as principal components)
are used for projection this can be written as:
$$X = 1\times{}\bar{x}^T + TP^T + V \tag{1}$$
Where the term $1\times{}\bar{x}^T$ represents the original variable
averages, $X$ denotes the observations, $T={t_1, t_2,\ldots,t_k}$ the
latent variables or scores, $P={p_1, p_2,\ldots,p_k}$
the transformation matrix (consisting of the most significant eigenvectors of the covariance matrix)
and $V$ are the residuals.

Missing values may be estimated by projecting the scores back into the original
space using $\hat{X} = 1\times{}\bar{x}^T + TP^T$.
Optimally, this produces an estimate of the missing data based on the
underlying correlation structure, thereby ignoring noise.
This will only produce reasonable results
if the residuals $V$ are sufficiently small, implying that most of the
important information is captured by the first $k$ components.

In order to calculate the transformation matrix $P$ one needs to determine
the covariance matrix between variables or alternatively calculate $P$
directly via SVD. In both cases, this can only be done on complete matrices.
However, an approximation may be obtained by use of different
regression methods.
The PCA methods provided in this package implement algorithms to accurately
estimate the PCA solution on incomplete data.

Although the focus of this package is clearly to provide a collection of
PCA methods we also provide a cluster based method for missing value imputation.
This allows to better rate and compare the results.

# Algorithms

All methods return a common class called `pcaRes` as a container
for the results. This guarantees maximum flexibility for the user. A wrapper
function called `pca()` is provided that receives the desired
type of pca as a string.

## svdPca

This is a wrapper function for R's standard `prcomp`
function. It delivers the results as a `pcaRes` object
for compatibility with the rest of the package.

## svdImpute

This implements the SVDimpute algorithm as proposed by Troyanskaya et al
[@troyanskaya01].
The idea behind the algorithm is to estimate the missing values as a
linear combination of the $k$ most significant eigengenes^[The
term "eigengenes" denotes the loadings when PCA was applied considering
variables (here the genes) as observations.].
The algorithm works iteratively until the change in the estimated solution
falls below a certain threshold.
Each step the eigengenes of the current estimate are calculated and
used to determine a new estimate.

An optimal linear combination is found by regressing an incomplete
variable against the $k$ most significant eigengenes. If the value at
position $j$ is missing,
the $j^{th}$ value of the eigengenes is not used when determining
the regression coefficients.

SVDimpute seems to be tolerant to relatively high amount of missing data
(> 10%).

## Probabilistic PCA (ppca)

Probabilistic PCA combines an EM approach for PCA with
a probabilistic model. The EM approach is based on the assumption that
the latent variables as well as the noise are normal distributed.

In standard PCA data which is far from the training set but close to the
principal subspace may have the same reconstruction error, see Figure
1 for explanation.

```{r pcaSubspacePrep}
library(pcaMethods)
x <- c(-4,7); y <- c(-3,4)
distX <- rnorm(100, sd=0.3)*3
distY <- rnorm(100, sd=0.3) + distX * 0.3
mat <- cbind(distX, distY)
res <- pca(mat, nPcs=2, method="svd", center=F)
loading <- loadings(res)[1,]
grad <- loading[2] / loading[1]
if (grad < 0)
   grad <- grad * -1
lx <- c(-4,7)
ly <- c(grad * -4, grad * 7)
```

```{r pcaSubspace, fig.width=8, fig.height=5, fig.cap="Normal distributed data with the first loading plotted in black. The two red points have the same reconstruction error because PCA does not define a density model. Thus the only measure of how well new data fits the model is the distance from the principal subspace. Data points far from the bulk of data but still close to the principal subspace will have a low reconstruction error."}
par(mar=c(2, 3, 2, 2))
plot(x,y, type="n", xlab="", ylab="")
abline(v=0, col="dark gray", lwd = 2); abline(h=0, col = "dark gray", lwd = 2)
points(distX, distY, type = 'p', col = "blue")
lines(lx,ly, lwd = 2)
points(-1, -1 * grad + 0.5, pch = 19, col = "red", lwd=4)
points(6, 6 * grad + 0.5, pch = 19, col = "red", lwd=4)
```

PPCA defines a likelihood function such that the likelihood for data
far from the training set is much lower, even if they are close to the
principal subspace.
This allows to improve the estimation accuracy.

PPCA is tolerant to amounts of missing values between 10% to 15%.
If more data is missing the algorithm is likely not to converge to
a reasonable solution.

The method was implemented after the draft
"EM Algorithms for PCA and Sensible PCA"
written by Sam Roweis and after the Matlab `ppca` script implemented
by *Jakob Verbeek*^[http://lear.inrialpes.fr/~verbeek/].

Please check also the PPCA help file.

## Bayesian PCA (bpca)

Similar to probabilistic PCA, Bayesian PCA uses an EM approach together
with a Bayesian model to calculate the likelihood for a reconstructed value.

The algorithm seems to be tolerant to relatively high amounts of missing data
(> 10%).
Scores and loadings obtained with Bayesian PCA slightly differ
from those obtained with conventional PCA.
This is because BPCA was developed especially for missing value estimation and
is based on a variational Bayesian
framework (VBF), with automatic relevance
determination (ARD). In BPCA, ARD leads to a different scaling of the
scores, loadings and eigenvalues when compared to standard PCA or
PPCA.
The algorithm does not force orthogonality between loadings.
However, the authors of the BPCA paper found that including an orthogonality criterion made the
predictions worse.
They also state that the difference between "real" and predicted
Eigenvalues becomes larger when the number of observation is smaller,
because it reflects the lack of information to accurately determine
true loadings from the limited and noisy data.
As a result, weights of factors to predict missing values are not the same as
with conventional PCA, but the missing value estimation is improved.

BPCA was proposed by Oba et al [@oba03].
The method available in this package is a port of the `bpca` Matlab
script also provided by the authors^[http://hawaii.aist-nara.ac.jp/%7Eshige-o/tools/].

## Inverse non-linear PCA (NLPCA)

NLPCA [@scholz05] is especially suitable for data from experiments
where the studied response is non-linear. Examples of such experiments
are ubiquitous in biology -- enzyme kinetics are inherently non-linear
as are gene expression responses influenced by the cell cycle or
diurnal oscillations.  NLPCA is based on training an auto-associative
neural network composed of a component layer which serves as the
"bottle-neck", a hidden non-linear layer and an output layer
corresponding to the reconstructed data.  The loadings can be seen as
hidden in the network.  Missing values in the training data are simply
ignored when calculating the error during back-propagation. Thus NLPCA
can be used to impute missing values in the same way as for
conventional PCA. The only difference is that the
loadings $P$ are now represented by a neural network.

A shortcoming of the current implementation is that there is no
reasonable stop criterion. The quality of the estimated solution
depends on the number of iterations. This should in most cases be
somewhat between 500 and 1500.  We recommend to use `kEstimate`
or `kEstimateFast` to determine this parameter.

## Nipals PCA

Nipals (Nonlinear Estimation by Iterative Partial Least Squares)
[@wold66] is an algorithm at the root of PLS regression which can
execute PCA with missing values by simply leaving those out from the
appropriate inner products.  It is tolerant to small amounts
(generally not more than 5%) of missing data.

## Local least squares (LLS) imputation

The package provides an algorithm called `llsImpute` for
missing value estimation based on a linear combination of the $k$
nearest neighbours of an incomplete variable (in Microarray
experiments normally a gene).  The distance between variables is
defined as the absolute value of the Pearson, Spearman or Kendall
correlation coefficient.  The optimal linear combination is found by
solving a local least squares problem as described in [@kim05]. In
tests performed in the cited paper the llsImpute algorithm is able to
outperform knnImpute [@troyanskaya01] and competes well with BPCA.

In the current implementation two slightly different ways for missing
value estimation are provided. The first one is to restrict the
neighbour searching to the subset of complete variables.  This is
preferable when the number of incomplete variables is relatively
small.

The second way is to consider all variables as candidates.
Here, missing values are initially replaced by the columns wise mean.
The method then iterates, using the current estimate as input for the
LLS regression until the change between new and old estimate falls below a
certain threshold (0.001).

# Getting started

**Installing the package.** To install the package first
download the appropriate file for your platform from the Bioconductor
website (http://www.bioconductor.org/). For Windows, start
`R` and select the `Packages` menu, then `Install
package from local zip file`.  Find and highlight the location of
the zip file and click on `open`.

For Linux/Unix, use the usual command `R CMD INSTALL` or set
the option `CRAN` to your nearest mirror site and use the
command `install.packages` from within an `R` session.

**Loading the package:** To load the `pcaMethods`
package in your `R` session, type `library(pcaMethods)`.

**Help files:** Detailed information on `pcaMethods`
package functions can be obtained from the help files.  For example,
to get a description of `bpca` type `help("bpca")`.

**Sample data:** Two sample data sets are coming with the
package. `metaboliteDataComplete` contains a complete subset
from a larger metabolite data set. `metaboliteData` is the same
data set but with 10 % values removed from an equal distribution.

# Some examples

```{r loadLibraries}
library(lattice)
library(pcaMethods)
```

To load the package and the two sample data sets type:

```{r loadData}
library(pcaMethods)
data(metaboliteData)
data(metaboliteDataComplete)
```

Now centre the data

```{r centerData}
md  <- prep(metaboliteData, scale="none", center=TRUE)
mdC  <- prep(metaboliteDataComplete, scale="none", center=TRUE)
```

Run SVD pca, PPCA, BPCA, SVDimpute and nipalsPCA on the data, using
the `pca()` wrapper function. The result is always a `pcaRes`
object.

```{r runPca}
resPCA  <- pca(mdC, method="svd", center=FALSE, nPcs=5)
resPPCA  <- pca(md, method="ppca", center=FALSE, nPcs=5)
resBPCA  <- pca(md, method="bpca", center=FALSE, nPcs=5)
resSVDI  <- pca(md, method="svdImpute", center=FALSE, nPcs=5)
resNipals  <- pca(md, method="nipals", center=FALSE, nPcs=5)
resNLPCA <- pca(md, method="nlpca", center=FALSE, nPcs=5, maxSteps=300)
```

Figure 2 shows a plot of the eigenvalue structure
(`sDev(pcaRes)`).
If most of the variance is captured
with few loadings PCA is likely to produce good missing value
estimation results.
For the sample data all methods show similar eigenvalues.
One can also see
that most of the variance is already captured by the first loading,
thus estimation is likely to work fine on this data.
For BPCA, the eigenvalues are scaled differently for reasons discussed above,
see Figure 3. The order of the loadings remains the
same.

```{r eigenvalues, fig.width=6, fig.height=4, fig.cap="Eigenvalue structure as obtained with different methods."}
sDevs <- cbind(sDev(resPCA), sDev(resPPCA), sDev(resBPCA), sDev(resSVDI), sDev(resNipals), sDev(resNLPCA))
matplot(sDevs, type = 'l', xlab="Eigenvalues", ylab="Standard deviation of PC", lwd=3)
legend(x="topright", legend=c("PCA", "PPCA", "BPCA", "SVDimpute","Nipals PCA","NLPCA"), lty=1:6, col=1:6, lwd=3)
```

To get an impression of the correctness of the estimation
it is a good idea to plot the scores / loadings obtained with classical
PCA and one of the probabilistic methods against each other. This of
course requires a complete data set from which data is randomly removed.
Figure 3 shows this for BPCA on the sample data.

```{r loadingBPCA, fig.width=7, fig.height=4, fig.cap="Loading 1 and 2 calculated with BPCA plotted against those calculated with standard PCA."}
par(mfrow=c(1,2))
plot(loadings(resBPCA)[,1], loadings(resPCA)[,1], xlab="BPCA", ylab="classic PCA", main = "Loading 1")
plot(loadings(resBPCA)[,2], loadings(resPCA)[,2], xlab="BPCA", ylab="classic PCA", main = "Loading 2")
```

# Cross validation

`Q2` is the goodness measure used for internal cross
validation. This allows to estimate the level of structure in a data
set and to optimise the choice of number of loadings.
Cross validation is performed by removing random elements of
the data matrix, then estimating these using the PCA algorithm of
choice and then calculating $Q^2$ accordingly. At the moment,
cross-validation can only be performed with algorithms that allow
missing values (i.e. not SVD). Missing value independent
cross-validation is scheduled for implementation in later versions.
$Q^2$ is defined as following for the mean centered data (and possibly
scaled) matrix $X$.

$$\mathrm{SSX}=\sum (x_{ij})^2$$
$$\mathrm{PRESS}=\sum (x_{ij} - \hat{x}_{ij})^2$$
$$Q^2=1 - \mathrm{PRESS}/\mathrm{SSX}$$

The maximum value for $Q^2$ is thus 1 which means that all variance in
$X$ is represented in the predictions; $X=\hat{X}$.

```{r computeQ2}
q2SVDI <- Q2(resSVDI, mdC, fold=10)
q2PPCA <- Q2(resPPCA, mdC, fold=10)
```

```{r fixQ2}
# PPCA does not converge / misestimate a value in very rare cases.
# This is a workaround to avoid that such a case will break the
# diagram displayed in the vignette.
# From the 2.0 release of bioconductor on, the convergence threshold
# for PPCA was lowert to 1e-5, this should make the method much more
# stable. So this workaround might be obsolete now...
# [nope it is not, ppca is unstable]
while( sum((abs(q2PPCA)) > 1) >= 1 ) {
    q2PPCA <- Q2(resPPCA, mdC, fold=10)
}
```

```{r Q2plot, fig.width=12, fig.height=7, fig.cap="Boxplot of the Q2 results for BPCA, Nipals PCA, SVDimpute and PPCA. PPCA and SVDimpute both deliver better results than BPCA and Nipals in this example."}
q2 <- data.frame(Q2=c(drop(q2PPCA), drop(q2SVDI)),
                 method=c("PPCA", "SVD-Impute")[gl(2, 5)], PC=rep(1:5, 2))
print(xyplot(Q2~PC|method, q2, ylab=expression(Q^2), type="h", lwd=4))
```

The second method called `kEstimate` uses cross validation to
estimate the optimal number of loadings for missing value
estimation.  The `NRMSEP` (normalised root mean square error of prediction)
[@feten05] or Q2 can be used to define the average error of prediction. The
NRMSEP normalises the square difference between real and estimated
values for a certain variable by the variance within this variable. The idea
behind this normalisation is that the error of prediction will
automatically be higher if the variance is higher.  The
`NRMSEP` for mean imputation is $\sqrt{\frac{nObs}{nObs - 1}}$
when cross validation is used, where $nObs$ is the number of
observations.  The exact definition is:
$$NRMSEP_k = \sqrt{\frac{1}{g} \sum_{j \in G} \frac{\sum_{i \in O_j}
(x_{ij} - \hat{x}_{ijk})^2}{o_j s_{x_j}^2}} \tag{2}$$
where $s^2_{x_j} = \sum_{i=1}^n (x_{ij} - \overline{x}_j)^2 / (n - 1)$,
this is the variance within a certain variable.
Further, $G$ denotes the set of incomplete variables, $g$ is the number of
incomplete varialbes. $O_j$ is the set of missing observations in
variable $j$ and $o_j$ is the number of missing observations in variable $j$.
$\hat{x}_{ijk}$ stands for the estimate of value $i$ of variable $j$ using $k$
loadings. See Figure 5 for an example.
The NRMSEP should be the error measure of choice. But if the number of
observations is small, the variance within a certain variable may become
and unstable criterion. If so or if variance scaling was applied we
recommend to use Q2 instead.

```{r kEstimateRun, echo=TRUE}
errEsti <- kEstimate(md, method = "ppca", evalPcs=1:5, nruncv=1, em="nrmsep")
```

```{r kEstimate, fig.width=4, fig.height=5, fig.cap="Boxplot showing the NRMSEP versus the number of loadings. In this example only 1 iteration of the whole cross validation were performed. It is normally advisable to do more than just one iteration."}
barplot(drop(errEsti$eError), xlab="Loadings", ylab="NRMSEP (Single iteration)")
```

`kEstimate` also provides information about the estimation error for individual variables.
The $Q^2$ distance or the NRMSEP are calculated separately for each variable.
See the manpage for `kEstimate` and `kEstimateFast` for details.
Plotting the variable - wise results gives information about for which variables
missing value estimation makes sense, and for which no imputation or mean imputation
is preferable, see Figure 6.
If you are not interested in variable - wise information we recommend to use the faster
`kEstimateFast` instead.

```{r variableWiseError, fig.width=5, fig.height=5, fig.cap="Boxplot showing the NRMSEP for all incomplete variables in the data set. For the first 7 variables missing value imputation does not seem to make too much sense."}
barplot(drop(errEsti$variableWiseError[, which(errEsti$evalPcs == errEsti$bestNPcs)]), xlab="Incomplete variable Index", ylab="NRMSEP")
```

# Visualisation of the results

## Quick scores and loadings plot

Some methods for display of scores and loadings are also provided.
The function `slplot()` aims to be a simple way to quickly
visualise scores and loadings in an intuitive way, see Figure
7. Barplots are provided when plotting only one PC and
colours can be specified differently for the scores and loadings
plots. For a more specific scatter plot it is however recommended to
access scores and loadings slots and define own plot functions.

```{r slplot, fig.width=11, fig.height=7, fig.cap="slplot for scores and loadings obtained with classical SVD based PCA."}
slplot(resPCA)
```

Another method called `plotPcs()` allows to visualise many
PCs plotted against each other, see Figure 8.

```{r plotPcs, fig.width=8, fig.height=8, fig.cap="A plot of score 1:3 for PPCA created with plotPcs()."}
plotPcs(resPPCA, pc=1:3, type="score")
```

## Using ggplot2

For using ggplot, the scores and loadings should best be added to a data frame that add other relevant descriptive factors. For example, after doing PCA on the Iris dataset, we may add the scores back to the original data frame and use ggplot to visualise, see Figure 9.

```{r ggplot, fig.width=4.5, fig.height=3, fig.cap="Score plot using ggplot2."}
pc <- pca(iris)
irdf <- merge(iris, scores(pc), by=0)
library(ggplot2)
ggplot(irdf, aes(PC1, PC2, colour=Species)) +
  geom_point() +
  stat_ellipse()
```
